High Resolution Identification of Protein-DNA Binding Events and Quality Control for ChIP-exo data

### 2015-11-20 rmarkdown/chip_exo_paper.tex
first draft of paper

%\VignetteEngine{knitr::knitr}
\documentclass{bmcart}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

Using the mouse-FoxA1 experiment from \cite{exoillumina}, the
relationship between this two quantities, library complexty and sample
enrichment was explored. Figure \ref{fig:enrich}A shows hexbin plots
illustrating the interaction between these two quantities. Notice how
there are two strong arms in each panel: The first one corresponds to
regions with low $\mbox{ARC}$ values and $\mbox{URCR}$ values that can
vary across the whole $[0,1]$ interval, while the second one shows a
decreasing trend in $\mbox{URCR}$ as $\mbox{ARC}$ increases. When an
experiment shows a higher degree of enrichment, then the separation of
this two arms is more noticeable, since the second arm corresponds to
possibly enriched regions (\ref{fig:enrich2}A and \ref{fig:enrich2}B).

Figure \ref{fig:enrich}B shows boxplots of the $\mbox{local-NSC}$ on
regions common to the 3 replicates and stratified by the number of
reads mapped exactly to one position for all $\sigma^{70}$ ChIP-exo
experiments; The high stratum is defined by regions consising of more
than 100 unique positions, the medium stratum for regions
where the number of unique position is on the
(50,100) range and the low stratum in the
(20,50) range. For each stratum and biological
replicate, we sampled 400 regions whenever possible (in
the opposite case, all the regions in the category were considered)
and calculated the $\mbox{local-NSC}$ for those regions. For all three
replicates, it is shown an increasing trend as the number of unique
position increases, which means that the $\mbox{local-NSC}$ is an
effective indicator of library complexity. Figure \ref{fig:enrich}C
shows the $\mbox{local-NSC}$ coefficient calculated for the same set
of regions, stratified by the number of unique positions in the
regions of ERR336942 using the same criteria as before. In the high
panel it is shown that the ERR336942 sample performs slightly better
than ERR336956 and both outperform ERR336935, this agrees with
\ref{fig:enrich}A, where only in the left-most panel doesn't seem to
be a separation between the low $\mbox{ARC}$ arm and the other.

@@

The strand imbalance assessment is based in the observation that the
enriched regions usually have a higher concentration of fragments,
therefore we examine the FSR (defined as the ratio of the number of
forward stranded reads divided by the total number of reads in a given
region) as the region with lower depth are being filtered out. This
indicator is of particular importance, since several methods rely on
the ``peak-pair'' assumption. In table \ref{tab:qcbase}, we calculated
the FSR and noticed that for all the samples, it's value is close to
0.5, which means that there are roughly the same amount of reads in
both strands. However, figure \ref{fig:comp}B shows that this
assumption don't hold in practice.

In order to asses the strand imbalance we create the following
visualization in figure \ref{fig:strand}. A) presents the FSR's
behavior as lower depth regions are being filtered out, while B) shows
which percentage of the regions are composed by fragments of both
strand or only one. Roughly speaking in a lower quality dataset there
is going to be a higher amount of single stranded-regions. To evaluate
this statement, for each replicate of the FoxA1 dataset from
\cite{exoillumina}, we divided the regions formed by at least
100 fragments into into the ones that overlap with any of the
sample's peaks. The p.value for ERR36935 is 0.1039, for
ERR336942 is 0.0064 and for ERR336956 is
0.0871. This shows that only the middle sample from figure
\ref{fig:strand} is the least imbalance sample from the the three,
while the other two still show a higher imbalance.

\begin{figure}[h!]
  \centering  
  \includegraphics[width = .8\textwidth,page = 3]{../figs/Carroll_mice_for_paper/Strand_imbalance.pdf} 
  \caption{Strand imbalance QC plots for the same data as in figure
    \ref{fig:enrich}. A) FSR distribution quantiles as the lower depth
    regions are being filtered out. B) Stacked histogram with the
    proportion of regions that are formed by two strands or only one.}
  \label{fig:strand}
\end{figure}

@@

We first compared various factors that could affect binding site
identification between ChIP-exo and ChIP-Seq data. in order to compare
distribution of signal and background between ChIP-exo and ChIP-Seq
data, we calculated ChIP tag counts across the genome by counting the
number of reads mapping to each of 150 non-overlapping
window after extending reads by 150 to their 3' end
directions. ChIP tag counts in ChIP-exo data were linearly related to
ChIP tag counts in ChIP-Seq data for the regions with high ChIP tag
counts (Figure \ref{fig:comp}A). This implies that signals for
potential binding sites are well reproducible between ChIP-exo and
ChIP-Seq data. On the other hand, there was clear difference in the
background distribution between them. In ChIP-Seq data reads were
almost uniformly distributed over background (non-binding) regions and
the ChIP tag counts in there regions were significantly larger than
zero. In contrast, in ChIP-exo data, there was larger variation in
ChIP tag counts among background regions and ChIP tag counts were much
lower in these regions compared to ChIP-Seq data. There were also
large proportion of regions without any read in ChIP-exo data. These
results indicate that background distribution of ChIP-exo data is less
homogeneous than that of ChIP-Seq data.

@@

Figure \ref{fig:reso_all} shows different comparisons among ChIP-exo,
PET ChIP-Seq and SET ChIP-Seq. A regulonDB annotation was considered
identified if the distance between it and dPeak binding site estimate
was at most of 20 bp. That way, the sensitivity is defined as
the proportion of regulonDB annotations identified in a peak. In
figure \ref{fig:reso_all}A can be seen that the sensitivity of all
protocols increases as the average distance between binding sites
does. Despite that when the binding events in a peak are closer to
each other both ChIP-exo and PET ChIP-Seq are comparable, as the
distance increases ChIP-exo identifies a higher proportion of the
regulonDB annotations; additionally SET ChIP-Seq is significantly less
sensible than both ChIP-exo and PET ChIP-Seq. In figure
\ref{fig:reso_all}B the distance between a regulonDB annotation to its
closest prediction is compared for ChIP-exo, PET and SET ChIP-Seq;
while the first two are comparable, both outperform SET
ChIP-Seq.

In figures \ref{fig:reso_all}C and \ref{fig:reso_all}D, we observe the
behavior of dPeak's estimated parameters for single end data (both
ChIP-exo and ChIP-Seq). The $\delta$ parameter measures the distance
from the 5' end of the fragment to its respective binding event and
the $\sigma$ parameter indicate the fragments distribution
variability around their respective binding sites (more information in
\cite{dpeak}). For both parameters, it can be seen that the ChIP-exo
estimates are smaller than the SET ChIP-Seq estimates in average,
which agrees with the fact that the sample reads are allocated more
tightly around the binding events in ChIP-exo data. Hence we can
conclude that the peaks shape is quite different between ChIP-exo and
SET ChIP-Seq.

@@

We updated the dPeak algorithm from \cite{dpeak} and showed that
ChIP-exo is comparable in resolution and sensitivity with PET ChIP-Seq
and outperforms SET ChIP-Seq. ChIP-exo compared with PET and SET
ChIP-Seq at fixed depth sample is able to identify more targets at a
lower resolution.

@@

We considered dPeak Chung et al. \cite{dpeak}, GEM Guo et
al. \cite{gem} and MACE Wang et al. \cite{mace} for ChIP-exo data
analysis. For the dPeak algorithm, we used the R package \emph{dPeak}
version 1.0.0, which is available from
\url{http://dongjunchung.github.io/dpeak/}. For the GEM algorithm we
used its Java implementation version 0.9, which is available from
\url{http://cgs.csail.mit.edu/gem/}. For the MACE algorithm, we used
its Python implementation version 1.0, which is available from
\url{http://dldcc-web.brc.bcm.edu/lilab/MACE/docs/html/}. For the
PEAKZILLA algorithm, we used its Python implementation, which is
available in \url{http://www.starklab.org/data/peakzilla/}. Candidate
regions for dPeak were identified for each replicate of ChIP-exo data
using the MOSAiCS algorithm Kuan et al. \cite{mosaics} (one sample
analysis using false discovery rate of 0.0001) implemented as an R
package \emph{mosaics} 2.4 (available from \emph{Bioconductor}
\url{https://www.bioconductor.org/packages/release/bioc/html/mosaics.html}). We
further filtered out candidate region with average ChIP tag count less
than 3,000 to avoid potential false positives based in exploratory
analysis. These regions were also explicitly provided to the GEM
algorithm as candidate regions. Default tuning parameters were used
during model fitting for all methods. We tried to use the CexoR
algorithm with its R implementation version 1.8.0 available from
\url{http://bioconductor.org/packages/release/bioc/html/CexoR.html},
but it returned an error everytime we used it.

@@

>>>>

% \VignetteEngine{knitr::knitr}
\documentclass{bmcart}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

Using the mouse-FoxA1 experiment from \cite{exoillumina} and the
relationship between this two quantities, library complexity and
sample enrichment was explored. Figure \ref{fig:enrich}A shows hexbin
plots illustrating the interaction between these two
quantities. Notice how there are two strong arms in each panel: The
first one corresponds to regions with low $\mbox{ARC}$ values and
varying $\mbox{URCR}$ values across the $(0,1]$ interval, while the
second one shows a decreasing trend in $\mbox{URCR}$ as $\mbox{ARC}$
increases. When an experiment shows a higher degree of enrichment,
then the separation of this two arms is more noticeable, since the
second arm corresponds to possibly enriched regions
(\ref{fig:enrich2}A and \ref{fig:enrich2}B).

Figure \ref{fig:enrich}B shows box plots of the $\mbox{local-NSC}$ on
regions common to the 3 replicates and stratified by the number of
reads mapped exactly to one position for all $\sigma^{70}$ ChIP-exo
experiments; The high stratum is defined by regions consisting of more
than 100 unique positions, the medium stratum for regions
where the number of unique position is on the
(50,100) range and the low stratum in the
(20,50) range. For each stratum and biological
replicate, we sampled 400 regions whenever possible (in
the opposite case, all the regions in the category were considered)
and calculated the $\mbox{local-NSC}$ for those regions. For all three
replicates, it is shown an increasing trend as the number of unique
position increases, which means that the $\mbox{local-NSC}$ is an
effective indicator of library complexity. Figure \ref{fig:enrich}C
shows the $\mbox{local-NSC}$ coefficient calculated for the same set
of regions, stratified by the number of unique positions in the
regions of ERR336942 using the same criteria as before. In the high
panel it is shown that the ERR336942 sample performs slightly better
than ERR336956 and both outperform ERR336935, this agrees with
\ref{fig:enrich}A, where only in the left-most panel doesn't seem to
be a separation between the low $\mbox{ARC}$ arm and the other.

\emph{Comment: I think I need to change the name of the FoxA1 samples
  to Rep1 to Rep3, but I wasn't sure about this. Therefore I used the
  name of the bam files}

@@

The strand imbalance assessment is based in the observation that the
enriched regions usually have a higher concentration of fragments,
therefore we examined the FSR (defined as the ratio of the number of
forward stranded reads divided by the total number of reads in a given
region) as the region with lower depth are being filtered out. This
indicator is of particular importance, since several methods rely on
the ``peak-pair'' assumption. In table \ref{tab:qcbase}, we calculated
the FSR and noticed that for all the samples, it's value is close to
0.5, which means that there are roughly the same amount of reads in
both strands. However, figure \ref{fig:comp}B shows that this value is
not representative of the sample locally, therefore the assumption
doesn't hold in practice.

In order to asses the strand imbalance we created the following
visualization presented in figure \ref{fig:strand}. A) shows the FSR's
behavior as lower depth regions are being filtered out, while B) shows
which percentage of the regions are composed by fragments of both
strands or only one (forward or backward). As the number of reads per
region's lower bound increases, the quantiles tend to approach the
median. To evaluate this statement, for each replicate of the FoxA1
dataset, we divided the regions formed by at least 100
fragments into into the ones that overlap with any of the sample's
peaks. The p.value for ERR36935 is 0.1039, for ERR336942 is
0.0064 and for ERR336956 is 0.0871. This shows
that only the middle sample from figure \ref{fig:strand} is the least
imbalance sample from the the three, while the other two still show a
higher imbalance at the 100 level.

\begin{figure}[h!]
  \centering  
  \includegraphics[width = .8\textwidth,page = 3]{../figs/Carroll_mice_for_paper/Strand_imbalance.pdf} 
  \caption{Strand imbalance QC plots for the same data as in figure
    \ref{fig:enrich}. A) FSR distribution quantiles as the lower depth
    regions are being filtered out, all quantiles approach to the
    median as the lower bound increases. B) Stacked histogram with the
    proportion of regions that are formed by two strands or only one,
    in a good sample the single-stranded regions are going to be
    filtered out quickly as in the middle row.}
  \label{fig:strand}
\end{figure}

@@

We first compared various factors that could affect binding site
identification between ChIP-exo and ChIP-Seq data. In order to compare
distribution of signal and background between ChIP-exo and ChIP-Seq
data, we calculated ChIP tag counts across the genome by counting the
number of reads mapping to each of 150 non-overlapping
window after extending reads by 150 to their 3' end
directions. ChIP tag counts in ChIP-exo data were linearly related to
ChIP tag counts in ChIP-Seq data for the regions with high ChIP tag
counts (Figure \ref{fig:comp}A). This implies that signals for
potential binding sites are well reproducible between ChIP-exo and
ChIP-Seq data. On the other hand, there was clear difference in the
background distribution between them. In ChIP-Seq data reads were
almost uniformly distributed over background (non-binding) regions and
the ChIP tag counts in there regions were significantly larger than
zero. In contrast, in ChIP-exo data, there was larger variation in
ChIP tag counts among background regions and ChIP tag counts were much
lower in these regions compared to ChIP-Seq data. There were also
large proportion of regions without any read in ChIP-exo data. These
results indicate that background distribution of ChIP-exo data is less
homogeneous than that of ChIP-Seq data.

@@

Figure \ref{fig:reso_all} shows different comparisons among ChIP-exo,
PET ChIP-Seq and SET ChIP-Seq. A RegulonDB annotation was considered
identified if the distance between it and dPeak binding site estimate
was at most of 20 bp. That way, the sensitivity is defined as
the proportion of RegulonDB annotations identified in a peak. In
figure \ref{fig:reso_all}A can be seen that the sensitivity of all
protocols increases as the average distance between binding sites
does. Despite that when the binding events in a peak are closer to
each other, both ChIP-exo and PET ChIP-Seq are comparable, as the
distance increases ChIP-exo identifies a higher proportion of the
RegulonDB annotations; additionally SET ChIP-Seq is significantly less
sensible than both ChIP-exo and PET ChIP-Seq. In figure
\ref{fig:reso_all}B the distance between a RegulonDB annotation to its
closest prediction is compared for ChIP-exo, PET and SET ChIP-Seq;
while the first two are comparable, both outperform SET ChIP-Seq.

In figures \ref{fig:reso_all}C and \ref{fig:reso_all}D, we observe the
behavior of dPeak's estimated parameters for single end data (both
ChIP-exo and ChIP-Seq). The $\delta$ parameter measures the distance
from the 5' end of the fragment to its respective binding event and
the $\sigma$ parameter indicates the fragments distribution
variability around their respective binding sites (more information in
\cite{dpeak}). For both parameters, it can be seen that the ChIP-exo
estimates are smaller than the SET ChIP-Seq estimates in average,
which agrees with the fact that the sample reads are allocated more
tightly around the binding events in ChIP-exo data. Hence we can
conclude that the peaks shape is quite different between ChIP-exo and
SET ChIP-Seq.

@@

We updated the dPeak algorithm from \cite{dpeak} and showed that
ChIP-exo is comparable in resolution and sensitivity with PET ChIP-Seq
and outperforms SET ChIP-Seq. ChIP-exo compared with PET and SET
ChIP-Seq at fixed depth sample is able to identify more targets at a
lower resolution. We compared dPeak with another algorithms to
estimate binding locations in ChIPexo data, dPeak is comparable to
MACE and outperforms GEM in resolution. dPeak provided a striking
balance in sensitivity, specificity and spatial resolution for
ChIP-exo analysis.

@@

We considered dPeak Chung et al. \cite{dpeak}, GEM Guo et
al. \cite{gem} and MACE Wang et al. \cite{mace} for ChIP-exo data
analysis. For the dPeak algorithm, we used the R package \emph{dPeak}
version 1.0.0, which is available from
\url{http://dongjunchung.github.io/dpeak/}. For the GEM algorithm we
used its Java implementation version 0.9, which is available from
\url{http://cgs.csail.mit.edu/gem/}. For the MACE algorithm, we used
its Python implementation version 1.0, which is available from
\url{http://dldcc-web.brc.bcm.edu/lilab/MACE/docs/html/}. For the
PEAKZILLA algorithm, we used its Python implementation, which is
available in \url{http://www.starklab.org/data/peakzilla/}. Candidate
regions for dPeak were identified for each replicate of ChIP-exo data
using the MOSAiCS algorithm Kuan et al. \cite{mosaics} (one sample
analysis using false discovery rate of 0.0001) implemented as an R
package \emph{mosaics} 2.4 (available from \emph{Bioconductor}
\url{https://www.bioconductor.org/packages/release/bioc/html/mosaics.html}). We
further filtered out candidate region with average ChIP tag count less
than 3,000 to avoid potential false positives based in exploratory
analysis. These regions were also explicitly provided to the GEM
algorithm as candidate regions. Default tuning parameters were used
during model fitting for all methods. We tried to use the CexoR
algorithm with its R implementation version 1.8.0 available from
\url{http://bioconductor.org/packages/release/bioc/html/CexoR.html},
but it returned an error every time we used it.

@@

% LocalWords:  ChIP exo URCR
